A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility

### 2019-08-20 paper.Rmd
Use data dictionary not codebook, close #96

1. README: A Human readable description of the data
2. Codebook: Human readable dictionary of data contents
3. License: How to use and share the data
4. Citation: How you want your data to be cited
5. Machine readable meta data
6. Raw data: The original/first data provided
7. Scripts: To clean raw data ready for analysis
8. Analysis ready data: Final data used in analysis

@@

1. README
2. Codebook
3. License
4. Citation
5. Analysis ready data

@@

## Codebook: Human readable dictionary of data contents {#codebook}

Codebooks provide human readable description of the data, providing context on the nature and structure of the data. This helps someone not familiar with the data understand, and use the data. At a minimum they should contain:

@@

**Variable names** should be short, descriptive names with no spaces or special characters. In the most common tabular data case, these correspond to column names, for example, "job_position", "faculty_level", and "years_at_company". **Variable labels** are longer descriptions of variables. For example "University Job Position", "University Faculty Position Level", and  "Number of Years Worked at University" [@McGill-codebook; @Arslan2019]. **Variable codes** apply to categorical (factor) variables, and are the values for their contents. For example, a variable could contain answers to a question, with values and codes like so: `0 = no, 1 = yes`, or for another variable indicating indoors or outdoors: `0 = inside` and `1 = outside`. These should be consistent across similar variables to avoid problems where `0 = yes` for one variable, and `1 = yes` in another. Date variables should have consistent formatting. For example, all date information could be in format "YYYY-MM-DD", and this should not be mixed with "YYYY-DD-MM". Date formatting should be described in the variable labels. **Missing data** are values that should have been observed, but were not. The code for missingness should be documented in the codebook, and should nominally be `NA`. If the reason for missingness is known it should be recorded. For example, censored data, patient drop out, or measurement error can have different values, such as "unknown", -99, or other value codes [@White2013; @Broman2017].

@@

- Explain formats like XML, as people might come into formats that wany XML.
- The power lies in the contrast to a csv codebook, as there is a defined, nested structure.
- Ontologies and domain specific metadata standards such as EML might also be worth mentioning.
- introduce the holy grail of the semantic web: JSON-LD

- readers just been advised to make a:
  - a csv codebook,
  - put that in the README
  - then make a JSON file. 
Is there a way to discuss strategies on harmonising all these formats?

@@

Tooling for producing data documentation information speeds up and simplifies the process of sharing data. Machine-readable metadata that can be indexed by google is created using the `dataspice` package [@dataspice]. To help create codebooks, we recommend the `codebook` package in R [@Arslan2019], which also generates machine readable metadata. Codebooks are implemented in other software such as STATA, which provides a "codebook" command. Data can be packaged up in a "data package" with DataPackageR [@DataPackageR], which provides tools to wrap up data into an R package, while providing helpers with MD5sum checks that track versioning. Note that is different to Frictionlessdata's tabular data package spec [@frictionlessdata-data-pkg].

@@

One example of a very laborious, long-term study is the 50-hectare plot on Barro Colorado Island in Panama. Researchers at the field station have censused every single tree in a pre-defined plot 7 times over the past 30 years. More than a million trees have been counted. The data and metadata however are hard to reach. To obtain the data, one must fill out a form and agree to terms and conditions which require approval and authorship prior to publication. The biggest challenge with this study is that the data are stored on a personal FTP server of one of the authors. While the data are available in CSV and binary Rdata formats, the data storage servers do not have any metadata, README files or codebooks. A separate, public archive hosts the metadata (https://repository.si.edu/handle/10088/20925) in a PDF file (https://repository.si.edu/bitstream/handle/10088/20925/RoutputFull.pdf?sequence=1&isAllowed=y) that describe the fields.

@@

3. **Provide a codebook** or glossary for each variable. Codebooks provide a free-form way to document the data, their structure and underlying details of how they were collected (https://www.emgo.nl/kc/codebook/). More importantly they provide a way for future consumers of the data to understand how the variables were named, how missing data were documented, along with any additional information that might impact how the data are interpreted.

>>>>

1. README: A Human readable description of the data
2. Data dictionary: Human readable dictionary of data contents
3. License: How to use and share the data
4. Citation: How you want your data to be cited
5. Machine readable meta data: Make your data searchable
6. Raw data: The original/first data provided
7. Scripts: To clean raw data ready for analysis
8. Analysis ready data: Final data used in analysis

@@

1. README
2. Data dictionary
3. License
4. Citation
5. Analysis ready data

@@

## Data dictionary: Human readable dictionary of data contents {#data-dictionary}

A data dictionary provide human readable description of the data, providing context on the nature and structure of the data. This helps someone not familiar with the data understand, and use the data. At a minimum they should contain:

@@

**Variable names** should be short, descriptive names with no spaces or special characters. In the most common tabular data case, these correspond to column names, for example, "job_position", "faculty_level", and "years_at_company". **Variable labels** are longer descriptions of variables. For example "University Job Position", "University Faculty Position Level", and  "Number of Years Worked at University" [@McGill-codebook; @Arslan2019]. **Variable codes** apply to categorical (factor) variables, and are the values for their contents. For example, a variable could contain answers to a question, with values and codes like so: `0 = no, 1 = yes`, or for another variable indicating indoors or outdoors: `0 = inside` and `1 = outside`. These should be consistent across similar variables to avoid problems where `0 = yes` for one variable, and `1 = yes` in another. Date variables should have consistent formatting. For example, all date information could be in format "YYYY-MM-DD", and this should not be mixed with "YYYY-DD-MM". Date formatting should be described in the variable labels. **Missing data** are values that should have been observed, but were not. The code for missingness should be documented in the data dictionary, and should nominally be `NA`. If the reason for missingness is known it should be recorded. For example, censored data, patient drop out, or measurement error can have different values, such as "unknown", -99, or other value codes [@White2013; @Broman2017].

@@

- Explain formats like XML, as people might come into formats that wany XML.
- The power lies in the contrast to a csv data dictionary, as there is a defined, nested structure.
- Ontologies and domain specific metadata standards such as EML might also be worth mentioning.
- introduce the holy grail of the semantic web: JSON-LD

- readers just been advised to make a:
  - a csv data dictionary,
  - put that in the README
  - then make a JSON file. 
Is there a way to discuss strategies on harmonising all these formats?

@@

Tooling for producing data documentation information speeds up and simplifies the process of sharing data. Machine-readable metadata that can be indexed by google is created using the `dataspice` package [@dataspice]. To help create a data dictionary, we recommend the `codebook` package in R [@Arslan2019], which also generates machine readable metadata. Data dictionaries are implemented in other software such as STATA, which provides a "codebook" command. Data can be packaged up in a "data package" with DataPackageR [@DataPackageR], which provides tools to wrap up data into an R package, while providing helpers with MD5sum checks that track versioning. Note that is different to Frictionlessdata's tabular data package spec [@frictionlessdata-data-pkg].

@@

One example of a very laborious, long-term study is the 50-hectare plot on Barro Colorado Island in Panama. Researchers at the field station have censused every single tree in a pre-defined plot 7 times over the past 30 years. More than a million trees have been counted. The data and metadata however are hard to reach. To obtain the data, one must fill out a form and agree to terms and conditions which require approval and authorship prior to publication. The biggest challenge with this study is that the data are stored on a personal FTP server of one of the authors. While the data are available in CSV and binary Rdata formats, the data storage servers do not have any metadata, README files or a data dictionary. A separate, public archive hosts the metadata (https://repository.si.edu/handle/10088/20925) in a PDF file (https://repository.si.edu/bitstream/handle/10088/20925/RoutputFull.pdf?sequence=1&isAllowed=y) that describe the fields.

@@

3. **Provide a data dictionary** or glossary for each variable. Data dictionaries provide a free-form way to document the data, their structure and underlying details of how they were collected (https://www.emgo.nl/kc/codebook/). More importantly they provide a way for future consumers of the data to understand how the variables were named, how missing data were documented, along with any additional information that might impact how the data are interpreted.
